\section{Conclusion}\label{conclusion}

In this paper we present insights into developers' processes and perceptions as they resolve merge conflicts.
Through semi-structured interviews and two surveys, we investigate the problems that merge conflicts pose, from the developers perspective.
We find that the majority of developers use a reactive process when monitoring for merge conflicts.
They rely mostly on their own knowledge of the code and on the complexity of the conflict when deciding on how to approach it.
While some developers defer the merge conflict resolution, it can lead to increased complexity, to the point that they might need to throw away changes and reimplement.
Finally, we find that when the merge conflict resolution fails, developers will either take the work offline, or collaborate to successfully resolve it. 

Providing insights into developers' processes and perceptions is critical, so that we can design tools that help resolve the issues facing developers.
We provide actionable implications for researchers, tool builders, and developers to harness the results of our study.
In future work, we plan to explore whether these factors, needs, and desired toolset improvements can be seamlessly merged into tools or techniques that assist developers' workflows.

%Merge conflicts interrupt development flow.
%Our work explores the human perceptions of merge conflict resolution in version control systems. We conducted an analysis of 10 interviews and confirmed our findings with a survey of 162 developers.
%
%We asked how developers approach merge conflicts and found that they identify 8 factors of importance for the initial evaluation of merge conflict difficulty. 
%We then asked which factors impact the difficulty of a merge conflict resolution and found 10 factors that impact the difficulty of the merge conflict resolution process. 
%These sets of factors serve as a way to inform managers and future research about merge-conflict-related difficulties of software developers.
%
%We also asked if developer tools met developer needs for merge conflict resolutions and found 6 improvements which developers would like to see in these tools. 
%In addition, we share specific tool needs from our interviews to give an idea of what people really struggle with in their merge conflict resolution workflows.